TIMES Ireland Model: Residential Sector

### 2021-05-21 manuscript.tex
Update 21 may 2021

%% main text
\section{Introduction}
\label{sec:Introduction}
\subsection{Climate Targets}
\label{Intro:Climate Targerts}
\par % Paragraph 1 - Intro
Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets. 
\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has increased to at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) which complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target. Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emission 2030 target would change from 30\% to 45\%. Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002. An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.

@@

\subsection{Ireland: Residential Sector Background}
\label{Intro:IreRSDBack}

\par
Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling is discussed later in this section and highlighted in Fig.\ref{fig:BERAge}. Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.  
In summary, the residential sector is highly reliant on fossil fuels as shown in Fig.\ref{fig:ResFuel}, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.

\begin{figure}[!htbp]
 \centering
 \includegraphics[scale=0.5]{Figures/ResidentialFuelUsage.png}
 \caption{Historical Residential Fuel Consumption}
 \label{fig:ResFuel}
\end{figure}
% NOTE THE Y-AXIS IS NOT CORRECT - DATA FROM SEAI ENERGY DATA PORTAL

The 2021 building regulations in 2021 require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, helps to better understand the housing stock. The BERs are required for new houses, rented houses, sold houses and renovated houses.

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}. BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, I will explain how I approached this problem in the literature review.

@@

@@

Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland} \cite{Harney2020DeterminingIreland}
In Ireland,  DH networks are a new, despite being used for decade across central Europe. CAP2019 plans to have approx. 0.4\% of heating provided by DH in 2030, where the  Research in Ireland h
IrDEA Up to 57\% feasible - same as Denmark DH
37% of the Irish heat demand to district heatingshows that the district heating option provides a solution that ismore efficient than an individual heating solution for 80% 2050 reduction 
potential for 37% district heating nationally (Connollyet al., 2013a).
HVO, solar may also play a part.

@@

``How should Ireland set GHG targets in the residential sector?” 
Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\subsection{Modelling Approach}
\label{LitR:Approach}
\subsection{Supply}
\label{LitR:Supply}
\subsection{Demand}
\label{LitR:Demand}
\subsection{Assumptions}
\label{LitR:Assumptions}
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
- LI,er.alThe nationwide survey identified existing technologies, age,
income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.

>>>>

%% main text
\section{Introduction}
\label{sec:Introduction}
\subsection{Climate Targets}

\par % Paragraph 1 - Intro
Under the Climate Action and Low Carbon Development (Amendment) Act 2021, the Irish government has proposed legally binding national targets to reduce Greenhouse Gas (GHG) emissions by 51\% in 2030 compared to 2018. The Act also proposes a long-term target to achieve a climate neutral economy or ``net-zero" by 2050 \cite{2021Climate2021}. These targets accounts for all GHGs - both Emission Trading System (ETS) emissions and non-ETS emissions, and sets out to develop sectoral five year carbon budgets to aid progress of the forementioned long-term targets. 
\par The EU's first Nationally Determined Contribution (NDC) which had a 40\% GHG reduction target (first NDC), has increased to at least a 55\% GHG reduction target (updated NDC) by 2030 compared to 1990 levels \cite{SUBMISSIONSTATES}. Ireland's Climate Action Plan 2019 (CAP2019) which complies with EU's first NDC, from which Irelands 30\% Non-ETS GHG reduction target in 2030 compared to 2005 levels was derived  \cite{DepartmentofCommunicationsClimateActionandEnvironment2019Climate2019}. Climate Action Plan 2021 will have to be more ambitious than its predecessor, as it must at least reflect EU's updated NDC target. Assuming the share of contribution of ETS and Ireland's Non-ETS remains the same, Ireland's non-ETS GHG emission 2030 target would change from 30\% to 45\%. Overall Ireland's national 2030 target would be about 15\% more ambitious than the EU's updated NDC.
There are also renewable energy targets to consider for transport (RES-T), heat (RES-H) and electricity (RES-E) under renewable directive (EU) 2018/2001, and energy efficiency targets under (EU) 2018/2002. An overview of Ireland's climate targets are shown in Fig.\ref{fig:IrelandClimatetargets}.

@@

\subsection{Ireland: Residential Sector}
\label{Intro:IreRSD}

@@

\subsubsection{Background}
\label{Intro:IreRSDBack}
Ireland’s distinctive decarbonisation pathway, is further separated from other EU member states when focusing on the residential sector. This is largely down to Ireland’s high reliance on fossil fuels, low building thermal efficiency, larger dwellings and rural settlement patterns. This is illustrated by the fact only 5.8\% of the population of Ireland were living in apartments in 2018, the lowest in EU-27 and well below the 25.3\% average \cite{EurostatDistributionSurveyilc_lvho01}, furthermore 71.4\% of the Irish population are deemed to be living in dwellings too large, where the EU-27 rate is 33\% \cite{EurostatShareSurveyilc_lvho50a}. The low thermal efficiency of Ireland's dwelling is discussed later in this section and highlighted in Fig.\ref{fig:BERAge}. Furthermore Irelands fossil fuel dependency is underlined by home heating oil accounting for the largest share of fuel usage in the residential sector at 37\% \cite{SustainableEnergyAuthorityofIreland2018EnergySector}, of which all is imported \cite{OCleirigh2020ENERGYReport}.

In the last 30 years, the extraction of indigenous peat has declined and consequently peat consumption has halved, in 2018 peat accounted for 12.3\% of residential GHG emissions. Residential coal GHG emissions has also declined to 6.9\% in 2018 \cite{EnergySEAI}. However, the decline in peat and coal has been offset by more significant growth in oil in the same period, where oil accounted for 34.8\% of residential GHG emissions in 2018 \cite{Howley2005ENERGY-RELATEDIreland}. Gas which is only available in some urban areas, accounted for 15.5\% of residential GHG emission in 2018 \cite{Howley2005ENERGY-RELATEDIreland}.  
In summary, the residential sector is highly reliant on fossil fuels, consequently Ireland currently has the lowest renewable heat of  among EU member states, at only 6.3\% \cite{Eurostat2021ShareSourcesnrg_ind_ren}.

The 2021 building regulations require a 70\% reduction of CO2 emissions in new dwellings compared to 2005 \cite{DepartmentofHousingGov.ieBuildings}, but since the 2008 global financial crisis there has been a considerable slowdown in the growth of housing stock. According to Central Statistics Office (CSO) approx. 85\% of residential buildings are 2005 or older. The BER database, which represents half of the dwelling stock, highlights the affect of construction year on the expected energy consumption, as shown in Fig.\ref{fig:BERAge}. The BERs labels are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). The BERs are developed from the Energy Performance of Buildings Directive 2002/91/ EC, which helps to better understand the housing stock. BERs are required for new houses, rented houses, sold houses and renovated houses.

Theoretically G-rated BERs consume more energy and consequently have higher energy bills to maintain a comfortable internal temperature, but in reality G-rated dwellings have low internal temperatures, which results in higher excessive winter mortality \cite{Clinch2000HousingMortality}. BERs do not take account of actual usage based on occupancy behaviour or occupancy density, nor does it take account of the size of the dwelling. The BER labels do therefore leave a gap between actual consumption and expected consumption, the literature review explains how this problem was tackled.

@@

Ireland has one of the lowest shares of District Heating (DH) in Europe at less than 1\%  \cite{Gartland2016AIreland} \cite{Harney2020DeterminingIreland}
In Ireland,  DH networks are a new, despite being used for decade across central Europe. CAP2019 plans to have approx. 0.4\% of heating provided by DH in 2030, where the  Research in Ireland h
IrDEA Up to 57\% feasible - same as Denmark DH
37\% Connollyet al., 2013a). of the Irish heat demand to district heatingshows that the district heating option provides a solution that ismore efficient than an individual heating solution for 80\% 2050 reduction

HVO, solar may also play a part.

@@

``How should Ireland set GHG targets in the residential sector?” 
Choosing the Pathways which Create the Least Burden
and Offer the Most Opportunity for Ireland
\subsection{Modelling Methodology or Approach}
\label{LitR:Model}
\subsection{Supply}
\label{LitR:Supply}
\subsection{Demand}
\label{LitR:Demand}
\subsection{Assumptions}
\label{LitR:Assumptions}
 % LIT REW INPUT DATA - EU \cite{Simoes2013TheProject} 
- LI,er.alThe nationwide survey identified existing technologies, age,
income, region, dwelling characteristics, and knowledge of eco- technology as the six most influential factors for determining homeowners' preferences for heating systems. Among
%the new A-rated houses behave distinctively different which much less demand on space heating and with pumps and fans energy consumption usage increasing, while appliance end use is increasing across the BER ratings. The number of appliances has increased rapidly and while energy efficient improves are  reducing the energy per appliance, it is overshadowed by the growth in electrical appliances. There retrofitting grant which  requires building to have a HLI indication of 2 or less before a heat pump grant can be installs. Ireland’s aim of retrofitting 0.5 million homes from the existing 1.7 million in the before 2030. The effect of these targets also stretches out across the construction sector. The savings from retrofits will decrease fossil fuel energy usage, while most houses 00\% would need to get heat pumps too, to fulfil 0.4 million existing houses getting heat pumps.  This energy efficiency first approach would allow for future technology choices, when current boiler are not at end of life.
